An Efficient Interpolation Technique for Jump Proposals in Reversible-Jump Markov Chain Monte Carlo Calculations

### 2010-12-14 paper.tex
More text on kD-trees and finding the neighborhood associated with a point.

@@

\will{Mention Weinberg direct integration (Ref.~\cite{Weinberg2009}).}

@@

One way to produce such a sequence of samples is via the Metropolis-Hastings algorithm, first proposed by Metropolis et al.~in 1953 \cite{Metropolis:1953}, and later generalized by Hastings:
\begin{enumerate}
  \item Given a current state $\vtheta^{(i)}$, propose the next state $\vtheta^p$ by drawing from a jump proposal distribution with probability $Q(\vtheta^{(i)} \to \vtheta^p)$.  
  \item Compute the probability of accepting the proposed jump as
\be
p_{\textnormal{accept}}  \equiv \min\Bigl(1,  \frac{p(\vtheta^p|d)}{p(\vtheta^{(i)}|d)} \frac{Q(\vtheta^p \to
        \vtheta^{(i)})}{Q(\vtheta^{(i)} \to \vtheta^p)} \Bigr).
\ee
\item Pick a uniform random number $\alpha \in [0,1]$.  If $\alpha<  p_{\textnormal{accept}}$, accept the proposed jump, setting $\vtheta^{(i+1)}=\vtheta^p$.  Otherwise, reject the jump, and remain at the same location in parameter space for the next step, $\vtheta^{(i+1)}=\vtheta^{(i)}$.
 \end{enumerate}

@@

A kD-tree is a binary, space-partitioning tree.  To partition a set of
points into a kD-tree, begin by placing them in a rectangular box that
contains all of parameter space.  Then proceed recursively:
\begin{enumerate}
\item If the given box contains exactly one point, stop; this is a
  leaf of the tree.  Otherwise:
\item Choose a dimension along which to divide the points.  Divide the
  points in half along this dimension (or nearly in half, if the
  number of points is odd), forming two sub-boxes.  The ``left''
  sub-box contains the half (or nearly half) of the points that have
  small coordinates along the chosen dimension; the ``right'' sub-box
  contains the half (or nearly half) of the points that have large
  coordinates along the chosen dimension.
\item Return to step 1 with each of the sub-boxes.
\end{enumerate}

@@

We have successfully employed the RJMCMC technique described above
when evaluating several alternative models for the distribution of the
masses of black holes in X-ray binaries \cite{Farr:2010}.  We
performed a Bayesian analysis of the mass distribution of stellar-mass
black holes using the observed masses of 15 low-mass X-ray binary
systems undergoing Roche lobe overflow and five high-mass, wind-fed
X-ray binary systems.  We considered ten different mass distribution
models: Gaussian, double Gaussian, power law, exponential decay, log
normal, and 1-, 2-, 3-, 4-, and 5-bin histograms.  Each model was
described by between two and six parameters.  Model selection using
RJMCMC with kD trees allowed us to determine that the mass
distribution of the low-mass systems is best fit by a power-law, while
the distribution of the combined sample is best fit by the exponential
model.  Based on the model selection, we were able to determine which
models would provide the best information about astrophysically
relevant parameters, such as the minimum black hole mass.  We were
also able to conclude that the low-mass subsample is not consistent
with being drawn from the distribution of the combined population
\cite{Farr:2010}.

>>>>

\newcommand{\order}[1]{\mathcal{O}\left( #1 \right)}

@@

\will{Mention Weinberg direct integration (Ref.~\cite{Weinberg2009}).
  Also \cite{Ascasibar2005} in the cosmological context (not really
  kD-tree, but close).}

@@

One way to produce such a sequence of samples is via the Metropolis-Hastings algorithm, first proposed by Metropolis et al.~in 1953 \cite{Metropolis:1953}, and later generalized by Hastings:
\begin{enumerate}
  \item Given a current state $\vtheta^{(i)}$, propose the next state $\vtheta^p$ by drawing from a jump proposal distribution with probability $Q(\vtheta^{(i)} \to \vtheta^p)$.  
  \item Compute the probability of accepting the proposed jump as
\bel{eq:p-accept}
p_{\textnormal{accept}}  \equiv \min\Bigl(1,  \frac{p(\vtheta^p|d)}{p(\vtheta^{(i)}|d)} \frac{Q(\vtheta^p \to
        \vtheta^{(i)})}{Q(\vtheta^{(i)} \to \vtheta^p)} \Bigr).
\ee
\item Pick a uniform random number $\alpha \in [0,1]$.  If $\alpha<  p_{\textnormal{accept}}$, accept the proposed jump, setting $\vtheta^{(i+1)}=\vtheta^p$.  Otherwise, reject the jump, and remain at the same location in parameter space for the next step, $\vtheta^{(i+1)}=\vtheta^{(i)}$.
 \end{enumerate}

@@

A kD-tree is a binary, space-partitioning tree.  To partition a set of
points into a kD-tree, begin by placing them in a rectangular box that
contains all of parameter space.  Then proceed recursively:
\begin{enumerate}
\item If the given box contains exactly one point, stop; this is a
  leaf of the tree.  Otherwise:
\item Choose a dimension along which to divide the points.  Divide the
  points in half along this dimension (or nearly in half, if the
  number of points is odd), forming two sub-boxes.  The ``left''
  sub-box contains the half (or nearly half) of the points that have
  small coordinates along the chosen dimension; the ``right'' sub-box
  contains the half (or nearly half) of the points that have large
  coordinates along the chosen dimension.
\item Return to Step 1 with each of the sub-boxes, storing the
  resulting trees as sub-trees of the current box.
\end{enumerate}
The key algorithmic step in the production of a kD-tree is finding the
median point along a given dimension in order to divide the points in
half in Step 2.  For $n$ points, this can be accomplished in
$\order{n}$ time (see, e.g., Ref.~\cite{Press2007}).  If there are $N$
points in total, there are $\order{\log N}$ levels in the tree; at
each level, $\order{N}$ points must be processed once in the
median-finding algorithm.  Tree construction thus costs $\order{N \log
  N}$ in time, and the tree consumes $\order{N \log N}$ space.  Box
boundaries for a kD-tree constructed around a point set that is
normally distributed around the origin in two dimensions are shown in
Figure \ref{fig:kD-tree}.

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\textwidth]{kdtree.png}
  \end{center}
  \caption{\label{fig:kD-tree} The neighborhoods from a kD-tree
    constructed around a set of points that are normally distributed
    about the origin in two dimensions.  As the points become denser
    around the origin, the typical neighborhood gets smaller.  The
    interpolated PDF within a box of volume $V_i$ is $1/(N V_i)$,
    where $N$ is the total number of points (and boxes).}
\end{figure}

In order to use the kD-tree interpolation in an MCMC, we must be able
to quickly find the neighborhood associated with a given point.  This
can be accomplished in $\order{\log N}$ time and constant space with
the following algorithm.  Given the point, $\vtheta_i$---which could
either be a point in the chain of MCMC samples if we are trying to
propose a jump, or a previously proposed jump if we are trying to
compute the jump probability distribution (see
Eq.~\ref{eq:p-accept})---and the tree, $T$:
\begin{enumerate}
\item If $T$ contains exactly one sample point, then its box is the
  associated neighborhood.  Otherwise:
\item The tree $T$ has two sub-trees.  If the point $\vtheta_i$ is
  contained in the ``left'' sub-tree, then return to Step 1,
  considering this sub-tree; otherwise return to Step 1, considering
  the ``right'' sub-tree.
\end{enumerate}

@@

We have successfully employed the RJMCMC technique described above
when evaluating several alternative models for the distribution of the
masses of black holes in X-ray binaries \cite{Farr2010}.  We
performed a Bayesian analysis of the mass distribution of stellar-mass
black holes using the observed masses of 15 low-mass X-ray binary
systems undergoing Roche lobe overflow and five high-mass, wind-fed
X-ray binary systems.  We considered ten different mass distribution
models: Gaussian, double Gaussian, power law, exponential decay, log
normal, and 1-, 2-, 3-, 4-, and 5-bin histograms.  Each model was
described by between two and six parameters.  Model selection using
RJMCMC with kD trees allowed us to determine that the mass
distribution of the low-mass systems is best fit by a power-law, while
the distribution of the combined sample is best fit by the exponential
model.  Based on the model selection, we were able to determine which
models would provide the best information about astrophysically
relevant parameters, such as the minimum black hole mass.  We were
also able to conclude that the low-mass subsample is not consistent
with being drawn from the distribution of the combined population
\cite{Farr2010}.
